---
title: "Analysis of a dilution dataset"
author: "David Huang, Craig Gin and Ben Callahan"
date: "03/06/2023"
output: html_document
---

The data set is obtained from [ControllingContaminants16S Github repository](https://github.com/lakarstens/ControllingContaminants16S) ([Karstens et al. 2019](https://journals.asm.org/doi/10.1128/msystems.00290-19)).

## Set up

```{r, message=FALSE, warning=FALSE}

# Install decontam package from the Github
# library(devtools)
# devtools::install_github("hczdavid/decontam")

# load the packages
pkgs <- c("tidyverse", "decontam", "RColorBrewer", "ggpubr", "caret", "phyloseq")
lapply(pkgs, require, character.only = TRUE)

# load the data
inpath  <- "rdata/"
outpath <- "output/"

load(paste0(inpath, "mockDilutionsPrep.RData"))

otu     <- as.data.frame(mock_ps@otu_table) %>% as.matrix
otusub  <- otu[, colSums(otu != 0) > 1]
otusubp <- sweep(otusub, 1, rowSums(otusub), "/")
conc    <- mock_ps@sam_data$DNA_conc
```

## Run dectonam, tcontam and tcontamL

```{r}
res_decontam  <- isContaminant(otusub, conc=conc, method='frequency') # default frequency method is "decontam"
res_tcontam   <- isContaminant(otusub, conc=conc, method='frequency', frequency.method = "tcontam")
res_tcontamL  <- isContaminant(otusub, conc=conc, method='frequency', frequency.method = "tcontamL")

allres        <- data.frame(seqID    = rownames(res_decontam),  
                            prev     = res_decontam$prev, 
                            decontam = res_decontam$p, 
                            tcontam  = res_tcontam$p, 
                            tcontamL = res_tcontamL$p)


allres$name      <- rownames(allres)
allres$Group     <- ifelse(allres$seqID %in% mock_taxa, "truesample", "contaminant")
allres$prevgroup <- cut(allres$prev, c(0, 2, 5, 9), labels=c("2", "3-5", "6-9"))
allres$reads     <- colSums(otusub)
```

## Distribution of the overall p-values or p-scores 

```{r, message=FALSE, warning=FALSE}
allresplot <- allres %>% gather(key = "method", value = "score", c(3:5))
allresplot$method <- factor(allresplot$method, levels = c("tcontam", "tcontamL", "decontam"))
# histogram
histo_p <- ggplot(allresplot, aes(x = score, fill=prevgroup)) + 
  geom_histogram() + labs(x = 'p-values (or scores)', y='Number ASVs', fill = "Prevalence") + 
  facet_wrap(~method)+
  theme_bw() + 
  theme(legend.position = "top", axis.title = element_text(size=15), axis.text = element_text(size = 12),
        strip.text = element_text(size = 15),
        legend.text = element_text(size = 12),
        legend.title  = element_text(size = 15))
histo_p
#ggsave(histo_p, filename = "hist_oral_new.png", width = 8, height = 4, dpi = 500)

```

## Comparison of the pattern for tcontam, tcontamL and decontam
```{r}
# overall pattern
dt_plotall <- ggplot(allres, aes(x = decontam, y = tcontam, col = Group)) + 
  geom_point(size = 1) + labs(color = "Group") + 
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray") + 
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray") + 
  geom_abline(intercept = 0, slope = 1,linetype = "dashed", color = "gray")+
  theme_bw() + 
  theme(legend.position = "top", axis.title = element_text(size=15), axis.text = element_text(size = 12),
        strip.text = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title  = element_text(size = 15)) +
  scale_color_manual(values = c(RColorBrewer::brewer.pal(5, "Set1")[c(5,4)], "gray90"))
dt_plotall

ttl_plotall <- ggplot(allres, aes(x = tcontamL, y = tcontam, col = Group)) + 
  geom_point(size = 1) + labs(color = "Group") + 
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray") + 
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "gray") + 
  geom_abline(intercept = 0, slope = 1,linetype = "dashed", color = "gray")+
  theme_bw() + 
  theme(legend.position = "top", axis.title = element_text(size=15), axis.text = element_text(size = 12),
        strip.text = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title  = element_text(size = 15)) +
  scale_color_manual(values = c(RColorBrewer::brewer.pal(5, "Set1")[c(5,4)], "gray90"))
ttl_plotall


# allpattern_oral <- ggarrange(histo_p, ggarrange(dt_plotall, ttl_plotall, labels = c("B", "C"), nrow = 1, widths = c(0.5,0.5), common.legend = T), labels = c("A", ""), nrow = 2, heights = c(0.5, 0.4))
# ggsave(allpattern_oral, filename = "allpattern_oral.png", width = 10, height = 8, dpi = 800)


```

## Comparison of the performance for tcontam, tcontamL and decontam at ASV level

```{r}

cutoff <- 0.05

allres$tcon_res  <- allres$tcontam < cutoff
allres$decon_res <- allres$decontam < cutoff
allres$tconL_res <- allres$tcontamL < cutoff

# confusion matrix 
table(allres$Group, allres$tcon_res)
table(allres$Group, allres$tconL_res)
table(allres$Group, allres$decon_res)

cutoff <- 0.1

allres$tcon_res  <- allres$tcontam < cutoff
allres$decon_res <- allres$decontam < cutoff
allres$tconL_res <- allres$tcontamL < cutoff

# confusion matrix 
table(allres$Group, allres$tcon_res)
table(allres$Group, allres$tconL_res)
table(allres$Group, allres$decon_res)


# F score
mythresh <- seq(0.01, 0.5, 0.01)
dF <- c()
pF <- c()
pFl <- c()

for(i in 1:length(mythresh)){
  
  myclass1 <- caret::confusionMatrix(as.factor(ifelse(allres$decontam > mythresh[i], "truesample", "contaminant")), as.factor(allres$Group), mode = "everything")
  myclass2 <- caret::confusionMatrix(as.factor(ifelse(allres$tcontam > mythresh[i], "truesample", "contaminant")), as.factor(allres$Group), mode = "everything")
  myclass3 <- caret::confusionMatrix(as.factor(ifelse(allres$tcontamL > mythresh[i], "truesample", "contaminant")), as.factor(allres$Group), mode = "everything")
 
  dF[i] <- myclass1$byClass["F1"]
  pF[i] <- myclass2$byClass["F1"]
  pFl[i] <- myclass3$byClass["F1"]
}

fres     <- data.frame(Threshold = mythresh, decontam = dF, tcomtam = pF, tcontamL  = pFl) 
fresplot <- fres %>% gather(key = "Method", value = "F1", 2:4)

f1score <- ggplot(fresplot, aes(x = Threshold, y = F1, group = Method, color = Method)) + 
  geom_point() + 
  geom_line() + 
  theme_bw() + ylab("F1 Score")+ #ylim(c(0.9, 1))+
  theme(legend.position = c(0.8,0.2),axis.text = element_text(size=12),
        axis.title = element_text(size = 15), legend.title = element_text(size = 12), legend.text =  element_text(size = 12) )
f1score
#ggsave(f1score, filename = "f1score3.png", width = 5, height = 5, dpi = 500)

```
## Comparison of the performance for tcontam, tcontamL and decontam at reads level

```{r}
allres_reads <- allres[, c("decontam", "tcontam", "tcontamL", "Group", "reads")]
allres_reads$Group <- ifelse(allres_reads$Group == "truesample", 0, 1)
reads <- allres$reads

readsdata <- c()
for(i in 1:nrow(allres_reads)){
  readsdata <- rbind(readsdata, matrix(t(allres_reads[i,])[,1], nrow = reads[i], ncol = 5, byrow = T))
}
colnames(readsdata) <- colnames(allres_reads)
readsdata <- as.data.frame(readsdata)
readsdata$Group <- ifelse(readsdata$Group == 0, "truesample", "contaminant")


# F score
dFreads <- c()
pFreads <- c()
pFlreads <- c()

readsdata$Group <- as.character(readsdata$Group)

for(i in 1:length(mythresh)){
  
  myclass1 <- caret::confusionMatrix(as.factor(ifelse(readsdata$decontam > mythresh[i], "truesample", "contaminant")), as.factor(readsdata$Group), mode = "everything")
  myclass2 <- caret::confusionMatrix(as.factor(ifelse(readsdata$tcontam > mythresh[i], "truesample", "contaminant")), as.factor(readsdata$Group), mode = "everything")
  myclass3 <- caret::confusionMatrix(as.factor(ifelse(readsdata$tcontamL > mythresh[i], "truesample", "contaminant")), as.factor(readsdata$Group), mode = "everything")
 
  dFreads[i] <- myclass1$byClass["F1"]
  pFreads[i] <- myclass2$byClass["F1"]
  pFlreads[i] <- myclass3$byClass["F1"]
}

fres_reads    <- data.frame(Threshold = mythresh, decontam = dFreads, tcomtam = pFreads, tcontamL  = pFlreads)
fresplotreads <- fres_reads %>% gather(key = "Method", value = "F1", 2:4)

f1score <- ggplot(fresplotreads, aes(x = Threshold, y = F1, group = Method, color = Method)) + 
  geom_point() + 
  geom_line() + 
  theme_bw() + ylab("F1 Score")+ 
  theme(legend.position = c(0.8,0.2),axis.text = element_text(size=12),
        axis.title = element_text(size = 15), legend.title = element_text(size = 12), legend.text =  element_text(size = 12) )
f1score
```


