---
title: "Simulation Analysis"
author: "David Huang, Craig Gin and Ben Callahan"
date: "03/06/2023"
output: html_document
---

The microbiome data is simulated based on the oral microbiome data from Davis et al. (2018) using Dirichlet-multinomial distribution. Please see the "simulationCode" folder to get the simulation results Rdata used in this Markdown file.

For the type one error analysis, we do not mix contaminant sequences to the true sample sequences, and perform 2000 iterations.

For the power comparison, we mixed 50 contaminant DNA sequences to 200 true sample sequences, and perform 500 iterations. We consider high-biomass (contaminant < 5%), low-biomass (contaminant 5% - 15%), verylow-biomass (contaminant 15% - 50%).

## Type one error

```{r, message=FALSE, warning=FALSE}
# load the packages
library(tidyverse)

# load the data
inpath  <- "rdata/"
outpath <- "output/"

# load the results
load(paste0(inpath, "simu_real_type1.Rdata"))

nseq <- 200

alpha <- c(0.01, 0.05)
type1error <- matrix(NA, nrow = 2, ncol = 3, dimnames = list(c(0.01, 0.05), c("decontam", "tcontam", "tcontamL")))
for(i in 1:length(alpha)){
  alltype1res<- sapply(alltype1, function(x){
  c(sum(x$decontam < alpha[i])/nseq, sum(x$tcontam < alpha[i])/nseq, sum(x$tcontamL_0.05 < alpha[i])/nseq)
}) %>% t
  type1error[i,] <- colMeans(alltype1res, na.rm = T)
}
type1error



# QQ plot
alltype1comb <- Reduce("rbind", alltype1)

par(mfrow = c(1, 2))
qqplot(runif(1e5), alltype1comb$tcontam, main = "tcontam")
abline(0,1, col = "red")

qqplot(runif(1e5), alltype1comb$tcontamL_0.05, main = "tcontam_L")
abline(0,1, col = "red")

```

## Power comparision 

```{r}
load(paste0(inpath, "simu_real_power_high_all_theta.Rdata"))
load(paste0(inpath, "simu_real_power_low_all_theta.Rdata"))
load(paste0(inpath, "simu_real_power_verylow_all_theta.Rdata"))

tl <- length(allpower_low_all)

allfpr_high <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))
alltpr_high <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))

allfpr_low <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))
alltpr_low <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))

allfpr_verylow <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))
alltpr_verylow <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))


for(jjj in 1:3){
  
  if(jjj == 1){allpower_temp <- allpower_low_all
  }else if(jjj == 2){
    allpower_temp <- allpower_high_all
  }else{
    allpower_temp <- allpower_verylow_all
  }

for(i in 1:tl){
  
  allpower <- allpower_temp[[i]]
  
  
  decontam <- sapply(allpower, function(x){
    c(sum(x$decontam[1:200] < 0.05)/200, sum(x$decontam[201:250] < 0.05)/50)
  }) %>% t
  
  
  
  tcontam <- sapply(allpower, function(x){
    c(sum(x$tcontam[1:200] < 0.05)/200, sum(x$tcontam[201:250] < 0.05)/50)
  }) %>% t
  
  
  
  tcontamL <- sapply(allpower, function(x){
    c(sum(x$tcontamL_0.05[1:200] < 0.05)/200, sum(x$tcontamL_0.05[201:250] < 0.05)/50)
  }) %>% t
  
  
  
 mytemp <-  cbind(colMeans(decontam),
  colMeans(tcontam),
  colMeans(tcontamL))
 
 
 if(jjj == 1){
   allfpr_low[i, ] <- mytemp[1,]
   alltpr_low[i, ] <- mytemp[2,]
 }else if(jjj ==2){
   
   allfpr_high[i, ] <- mytemp[1,]
   alltpr_high[i, ] <- mytemp[2,]
 }else{
   allfpr_verylow[i, ] <- mytemp[1,]
   alltpr_verylow[i, ] <- mytemp[2,]
   
 }
 
}

}


theta_low     <- seq(0.0003, 0.0016, 0.0002)
theta_high    <- seq(0.00002, 0.00015, 0.00002)
theta_verylow <- seq(0.001, 0.004, 0.0005)


# false positive rate
allfpr_high_plot <- allfpr_high %>% as.data.frame(alltpr_high) %>% 
  mutate(theta = theta_high, Type = "High_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

allfpr_low_plot <- allfpr_low %>% as.data.frame(alltpr_low) %>% 
  mutate(theta = theta_low, Type = "Low_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

allfpr_verylow_plot <- allfpr_verylow %>% as.data.frame(alltpr_verylow) %>% 
  mutate(theta = theta_verylow, Type = "Verylow_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

allplot_fpr <- rbind(allfpr_high_plot, allfpr_low_plot, allfpr_verylow_plot)


myplot_fpr <- ggplot(allplot_fpr, aes(x = theta, y = TPR, group = Method, color = Method)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Type, scales = "free_x") + 
  theme_bw() + ylab("False positive rate") + xlab (expression(theta ~" (overdispersion parameter)"))
myplot_fpr

alltpr_high_plot <- alltpr_high %>% as.data.frame(alltpr_high) %>% 
  mutate(theta = theta_high, Type = "High_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

alltpr_low_plot <- alltpr_low %>% as.data.frame(alltpr_low) %>% 
  mutate(theta = theta_low, Type = "Low_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

alltpr_verylow_plot <- alltpr_verylow %>% as.data.frame(alltpr_verylow) %>% 
  mutate(theta = theta_verylow, Type = "Verylow_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

allplot_tpr <- rbind(alltpr_high_plot, alltpr_low_plot, alltpr_verylow_plot)

myplot_tpr <- ggplot(allplot_tpr, aes(x = theta, y = TPR, group = Method, color = Method)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Type, scales = "free_x") + 
  theme_bw() + ylab("False positive rate") + xlab (expression(theta ~" (overdispersion parameter)"))
myplot_tpr

#ggsave(myplot, file = "C:/Users/David Huang/Desktop/simutcontam_fpr_slide.png", width = 6, height = 3, dpi = 600)


```

## Power comparison with fixed theta

```{r}
load(paste0(inpath, "simu_real_power_all_theta.Rdata"))

allpower_high_all <- allpower[[1]]
allpower_low_all <- allpower[[2]]
allpower_verylow_all <- allpower[[3]]

tl <- length(allpower_low_all)

allfpr_high <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))
alltpr_high <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))

allfpr_low <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))
alltpr_low <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))

allfpr_verylow <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))
alltpr_verylow <- matrix(NA, nrow = tl, ncol= 3, dimnames = list(1:tl, c("decontam", "tcontam", "tcontamL")))


for(jjj in 1:3){
  
  if(jjj == 1){allpower_temp <- allpower_low_all
  }else if(jjj == 2){
    allpower_temp <- allpower_high_all
  }else{
    allpower_temp <- allpower_verylow_all
  }

for(i in 1:tl){
  
  allpower <- allpower_temp[[i]]
  
  
  decontam <- sapply(allpower, function(x){
    c(sum(x$decontam[1:200] < 0.05)/200, sum(x$decontam[201:250] < 0.05)/50)
  }) %>% t
  
  
  
  tcontam <- sapply(allpower, function(x){
    c(sum(x$tcontam[1:200] < 0.05)/200, sum(x$tcontam[201:250] < 0.05)/50)
  }) %>% t
  
  
  
  tcontamL <- sapply(allpower, function(x){
    c(sum(x$tcontamL_0.05[1:200] < 0.05)/200, sum(x$tcontamL_0.05[201:250] < 0.05)/50)
  }) %>% t
  
  
  
 mytemp <-  cbind(colMeans(decontam),
  colMeans(tcontam),
  colMeans(tcontamL))
 
 
 if(jjj == 1){
   allfpr_low[i, ] <- mytemp[1,]
   alltpr_low[i, ] <- mytemp[2,]
 }else if(jjj ==2){
   
   allfpr_high[i, ] <- mytemp[1,]
   alltpr_high[i, ] <- mytemp[2,]
 }else{
   allfpr_verylow[i, ] <- mytemp[1,]
   alltpr_verylow[i, ] <- mytemp[2,]
   
 }
 
}

}


theta_low     <- seq(0.001, 0.004, 0.001)
theta_high    <- seq(0.001, 0.004, 0.001)
theta_verylow <-seq(0.001, 0.004, 0.001)

# false positive rate
allfpr_high_plot <- allfpr_high %>% as.data.frame(alltpr_high) %>% 
  mutate(theta = theta_high, Type = "High_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

allfpr_low_plot <- allfpr_low %>% as.data.frame(alltpr_low) %>% 
  mutate(theta = theta_low, Type = "Low_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

allfpr_verylow_plot <- allfpr_verylow %>% as.data.frame(alltpr_verylow) %>% 
  mutate(theta = theta_verylow, Type = "Verylow_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

allplot_fpr <- rbind(allfpr_high_plot, allfpr_low_plot, allfpr_verylow_plot)


myplot_fpr <- ggplot(allplot_fpr, aes(x = theta, y = TPR, group = Method, color = Method)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Type, scales = "free_x") + 
  theme_bw() + ylab("False positive rate") + xlab (expression(theta ~" (overdispersion parameter)"))
myplot_fpr

alltpr_high_plot <- alltpr_high %>% as.data.frame(alltpr_high) %>% 
  mutate(theta = theta_high, Type = "High_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

alltpr_low_plot <- alltpr_low %>% as.data.frame(alltpr_low) %>% 
  mutate(theta = theta_low, Type = "Low_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

alltpr_verylow_plot <- alltpr_verylow %>% as.data.frame(alltpr_verylow) %>% 
  mutate(theta = theta_verylow, Type = "Verylow_Biomass") %>% 
  gather(key = "Method", value = "TPR", 1:3)

allplot_tpr <- rbind(alltpr_high_plot, alltpr_low_plot, alltpr_verylow_plot)

myplot_tpr <- ggplot(allplot_tpr, aes(x = theta, y = TPR, group = Method, color = Method)) + 
  geom_point() + 
  geom_line() + 
  facet_wrap(~Type, scales = "free_y") + 
  theme_bw() + ylab("False positive rate") + xlab (expression(theta ~" (overdispersion parameter)"))
myplot_tpr

```
